 

### 2021-01-26 vertical-thermal-review.Rmd
taking stock of/ organizing/ updating figures, including auto numbering
#28 , #29

# This chunk is for auto numbering of tables/figures

#figures
fig_schematic = "1"
fig_NEON_vertical =  "2"
fig_NEON_differences  = "3"

@@

#Appendices
appendix_NEON_methods <- "S1"
appendix_tealeaves_methods <- "S2"
appendix_lit_review_methods <- "S3"
appendix_SCBI_treering <- "S4"

appendix_NEON_methods_title<- "Methods for NEON vertical profiles"
appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
appendix_lit_review_methods_title<- "Methods for literature review"
appendix_SCBI_treering_title <- "Methods for SCBI tree-ring analysis"

@@

**We lack a systematic, up to date understanding of biophysical and biological patterns across the vertical canopy gradient in forests, how this gradient affects leaf-level processes, and in turn how it affects ecosystems (Fig. ` r fig_schematic`).**

@@

![**Figure 1. DRAFT schematic of a forest summarizing most important gradients--sort of a "graphical abstract".** . ](schematic/vertical thermal sensitivity image.png)

@@

*(paragraph on leaf area distribution, including figure ([separate from or combined with current Fig. 2- see GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
The vertical structure and corresponding microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019) and following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.

@@

During the wet season, maximum air temperatures have been shown to be higher above forest canopies than below in tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al. 2017]. Similarly, in these studies, maximum temperatures were higher while humidity was lower in gaps compared to understory regions (both at 1.5 m above the ground, see Fig. 4 - again illustrating buffering effect of the canopy). However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama, partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016]. Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].

@@

![**Figure 2. Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is just one of several+ intended sites. See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2. ](NEON_height_profiles/figures/profile_SCBI.png)

@@

![**Figure 3. Role of canopy permeability in shaping temperature buffering, from NEON data.** Current placeholder figure is Nidhi's first rough analysis. See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2. ](NEON_height_profiles/figures/differences_across_forests_temp.png)

@@

Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1. summarizes relevant traits for thermal sensitivity.

@@

**Many traits vary with height and/or between sun and shade leaves of the same species (Table 1).**
Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.

![(*For latest version of table, see tables.pdf.*)](table1.png)

@@

**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. 3).
Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020] 
$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [@michaletzPlantThermoregulationEnergetics2015].

@@

Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm). Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance. However, counteracting this, lower wind speed in understory would reduce latent heat loss. Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance. However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequentlly increasing leaf temperature in upper canopy [@fauset_differences_2018]. Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.

![**Figure 3. Leaf temperature in response to (a) XXX, (b) XXX (key trait or environmental variables).** Examples to be produced based on energy balance model using the tealeaves R package of @muir_tealeaves_2019: https://github.com/EcoClimLab/vertical-thermal-review/issues/6. Current placeholder shows figures from Campbell&Norman.](leaf_energy_balance/C&N_figures.png)

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. **2?**).**
[Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].]
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999). 
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).

@@

Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. 1). 
However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves (Table 2).

![(*For latest version of table, see tables.pdf.*)](table2.png)
```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below (easiest in alphabetical order)
```
---
nocite: | 
  @bolstad_foliar_1999; @roberts_stomatal_1990; @slot_photosynthetic_2019; @harleyEffectsLightTemperature1996; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @ishiiExploringRelationshipsCanopy2004; @curtis_intracanopy_2019; @bolstad_foliar_1999; @kafutiFoliarWoodTraits2020; 
---

@@

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ (Rollinson et al., in review).
Similarly, our previously unpublished analysis shows that in a broadleaf deciduous forest in Virginia, understory trees in a Mid-Atlantic forest responded more negatively to warmer $T_{air}$ than did canopy trees (Fig. **4**; Appendix `r appendix_SCBI_treering`). 
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. **3?**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. **3**).
It is also possible that competition is accelerated under warmer temperatures.

Additionally, increased exposure to light increases non structural carbon reserve (NSC), plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availbility become limited, however with available water overall sensitivity to stress decreases. Understory shaded/younger trees have comparitively less NSC reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotransition demands, as well as greater investment in light harvesting than in photosynthesis--with greater overall sensitivity to stress and reduced tolerance. Larger NSC reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature]. Thus, overall stress sensitivity decreases and tolerance increases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]

@@

Interaction between degrees of shade x ontogeny x height with rising temperature needs more scietific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.

![**Figure 4. Temperature sensitivity of tree growth (derived from tree rings) for understory versus canopy trees.** THIS IS A ROUGH PLACEHOLDER-- KAT WILL PLAY WITH POTENTIAL VERSIONS. IF OUR IN-PREP TREE RING ANALYSIS METHODS PAPER IS ACCEPTED AHEAD OF THIS, WE CAN HIGHLIGHT RESULTS FROM THAT, WHICH ARE MORE RELIABLE. Analysis methods described in Appendix S4. OUTLIER HERE IS SHADE-TOLERANT BEECH. Data from Helcoski et al. 2019.](SCBI_tree_rings/SCBIfig.png)

@@

Appendix `r appendix_SCBI_treering`. `r appendix_SCBI_treering_title`

>>>>

# DISPLAY ITEMS  (limit 8) # This chunk is for auto numbering of tables/figures
#figures
fig_schematic = 1
fig_NEON_vertical =  2
fig_NEON_differences  = 3
fig_leaf_T = 4
fig_treerings= 5 #not sure if we want to include one

#tables
table_leaf_traits=1
table_leaf_metabolism=2

@@

#Appendices
appendix_NEON_methods <- "S1"
appendix_tealeaves_methods <- "S2"
appendix_lit_review_methods <- "S3"

appendix_NEON_methods_title<- "Methods for NEON vertical profiles"
appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
appendix_lit_review_methods_title<- "Methods for literature review"

@@

**We lack a systematic, up to date understanding of biophysical and biological patterns across the vertical canopy gradient in forests, how this gradient affects leaf-level processes, and in turn how it affects ecosystems (Fig. `r fig_schematic`).**

@@

![**Figure `r fig_schematic`. DRAFT schematic of a forest summarizing most important gradients--sort of a "graphical abstract".** [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27). ](schematic/vertical thermal sensitivity image.png)

@@

*(paragraph on leaf area distribution, including figure ([GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
The vertical structure and corresponding microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019) and following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.

@@

During the wet season, maximum air temperatures have been shown to be higher above forest canopies than below in tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al. 2017]. Similarly, in these studies, maximum temperatures were higher while humidity was lower in gaps compared to understory regions (both at 1.5 m above the ground, see their Fig. 4 - again illustrating buffering effect of the canopy). However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama, partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016]. Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].

@@

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is just one of several+ intended sites. See [issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2) and [issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20). ](NEON_height_profiles/figures/profile_all.png)

@@

![**Figure `r fig_NEON_differences`. Role of canopy permeability in shaping temperature buffering, from NEON data.** Current placeholder figure is Nidhi's first rough analysis. See [issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2). ](NEON_height_profiles/figures/differences_across_forests_temp.png)

@@

Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

@@

**Many traits vary with height and/or between sun and shade leaves of the same species (Table `r table_leaf_traits`).**
Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.

![Table `r table_leaf_traits`. (*For latest version of table, see tables.pdf.*)](table1.png)

@@

**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`).
Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020] 
$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [@michaletzPlantThermoregulationEnergetics2015].

@@

Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm). Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance. However, counteracting this, lower wind speed in understory would reduce latent heat loss. Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance. However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018]. Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.

![**Figure `r fig_leaf_T`. Leaf temperature in response to (a) XXX, (b) XXX (key trait or environmental variables).** Examples to be produced based on energy balance model using the tealeaves R package of @muir_tealeaves_2019: https://github.com/EcoClimLab/vertical-thermal-review/issues/6. Current placeholder shows figures from Campbell&Norman.](leaf_energy_balance/C&N_figures.png)

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).**
[Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].]
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999). 
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).

@@

Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves (Table `r table_leaf_metabolism`).

![Table `r table_leaf_metabolism`. (*For latest version of table, see tables.pdf.*)](table2.png)
```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below (easiest in alphabetical order)
```
---
nocite: | 
  @bolstad_foliar_1999; @roberts_stomatal_1990; @slot_photosynthetic_2019; @harleyEffectsLightTemperature1996; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @ishiiExploringRelationshipsCanopy2004; @curtis_intracanopy_2019; @bolstad_foliar_1999; @kafutiFoliarWoodTraits2020; 
---

@@

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; Rollinson et al., in press].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).
It is also possible that competition is accelerated under warmer temperatures.

Additionally, increased exposure to light increases non structural carbon reserve (NSC), plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availability become limited, however with available water overall sensitivity to stress decreases. Understory shaded/younger trees have comparatively less NSC reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotranpiration demands, as well as greater investment in light harvesting than in photosynthesis--with greater overall sensitivity to stress and reduced tolerance. Larger NSC reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature]. Thus, overall stress sensitivity decreases and tolerance increases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]

@@

Interaction between degrees of shade x ontogeny x height with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.

![**Figure `r fig_treerings`. DRAFT tree-rings figure. Current figure shows (a) drought responses as a function of height in a temperate broadleaf forest (SCBI), from McGregor et al. 2020; (b) responses to growing season temperature in eight New England forests, from Rollinson et al. 2020.  I'M NOT SURE THIS IS WHAT WE WANT, BUT I THINK SOME SORT OF TREE-RING FIGURE WOULD BE GOOD. [ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/tree_rings_draft.png)

@@
